---
title: "Time Well Spent? A Study on How High School Students’ Activities Affect Their Grades"
subtitle: "ECON 471 Research Project"
author: "Ramsey El Lethy, Darwin Mah"
date : 04-29-2024
header-includes:
    - \usepackage{setspace}\doublespacing
    - \usepackage{placeins}
output: pdf_document
---

## Introduction 

It is no secret that, as students, we see a direct relationship between the time we invest in studying concepts and our exam performance. However, students' lives incorporate so much more activity than just studying. Whether finding time to care for the people in our lives or setting aside a moment for ourselves, students consistently add numerous activities to their schedules. Some of these are just too much, some seriously help, and some have no effect. Time is significant to us, so learning how we spend it can give us insights into our opportunities and shortcomings as we traverse this educational journey. 

## Abstract
To assess the impact of how students divide their time on their academic performance, we acquired data from two thousand individuals, all of whom are high schoolers of different genders and ages. Within the data set, each student was recorded, including whether they took part in extracurricular activities if they had a part-time job, the number of absences, and the number of hours students spent studying a week. We also recorded test scores for multiple subjects, including math, history, physics, chemistry, biology, english, and geography. Our model represented true or false records of having a part-time job or an activity as a 0 for false and a 1 for true. We performed multiple regressions, each for every individual subject. Regressing the same independent variables on the different scores. We aimed to identify the relationship between these factors and their test scores. 

\newpage
## Observations and Analysis 
We perform multiple regression analyses for multiple subjects to find consistent relationships between the variables across each 
 subject. The first thing we looked at was the individual significance of each 
 variable; we found that consistently, the only insignificant factor was
 whether or not students participated in extracurricular activities. In our 
 analysis of the overall significance of the model, a rejection of the null
 hypothesis due to a large F statistic coincides with at least 
 one of our parameters, in this case, extracurricular, not being able to 
 explain student performance in any capacity. We see consistent relationships between student 
 performance for the rest of our parameters. Table 1 below presents the values of the 
 coefficients and respective tests. Aligning with common perceptions about
 studying, we see that our weekly self-study hour parameter is positive in 
 all models. This means that an increase in time spent studying increases student
 performance. Looking at students' absences, we see a negative
 relationship between being at school and their performance. Interestingly, 
 students with a job have decreased math, 
 physics, and history performance. However, biology gave us conflicting results, expressing a 
 positive relationship between having a job and performance.



```{r Regression Analysis, results = 'asis', echo = FALSE, message = FALSE }
student_data <- read.csv("student-scores.csv", header = TRUE)

#Representing our dummy vairables as numeric values
extra_curricular <- as.numeric(student_data$extracurricular_activities == "True")
has_job <- as.numeric(student_data$part_time_job == "True")

#Regressive Stastics 
student_regression_math <- 
  lm(math_score ~ absence_days + 
  weekly_self_study_hours +extra_curricular+has_job, data = student_data)

student_regression_phys <- 
  lm(physics_score ~ absence_days + 
  weekly_self_study_hours +extra_curricular+has_job, data = student_data)

student_regression_his <- 
  lm(history_score ~ absence_days + 
  weekly_self_study_hours +extra_curricular+has_job, data =student_data)

student_regression_bio <- 
  lm(biology_score ~ absence_days + 
  weekly_self_study_hours +extra_curricular + has_job, data =student_data)


library(dplyr)
library(knitr)
library(kableExtra)

# Function to create a data frame of coefficients for a given model
create_coefficients_dataframe <- function(regression_model, model_name) {
  # Extract the summary of the model
  model_summary <- summary(regression_model)
  
  # Extract coefficients and their statistics
  coefficients <- model_summary$coefficients  # This includes Estimate, Std. Error, t value, and Pr(>|t|)
  
  # Extract F-statistic and its degrees of freedom
  f_statistic <- model_summary$fstatistic["value"]
   df1 = model_summary$fstatistic["numdf"]  # Degrees of freedom for the model
  df2 = model_summary$fstatistic["dendf"]  # Degrees of freedom for the residuals
  f_p_value = pf(f_statistic, df1, df2, lower.tail = FALSE)  # Calculating the p-value for the F-statistic
  
  # Create a data frame containing the coefficients and additional model details
  df <- data.frame(
    Model = model_name,
    CoefficientValue = coefficients[, "Estimate"],
    StdError = coefficients[, "Std. Error"],
    TValue = coefficients[, "t value"],
    Pr = coefficients[, "Pr(>|t|)"],
    FStatistic = rep(f_statistic, nrow(coefficients)),
    ResidualDF = rep(df2, nrow(coefficients))
  )
  return(df)
}

# Example of applying the function to a model
df_math <- create_coefficients_dataframe(student_regression_math, "Math")
df_phys <- create_coefficients_dataframe(student_regression_phys, "Physics")
df_his <- create_coefficients_dataframe(student_regression_his, "History")
df_bio <- create_coefficients_dataframe(student_regression_bio, "Biology")

# Combine all data frames into one
all_coefficients_df <- dplyr::bind_rows(df_math, df_phys, df_his, df_bio)



# Display the table with enhanced formatting for PDF
 all_coefficients_df %>%
  kable("latex", booktabs = TRUE, caption = "Summary Statistics for Original Model") %>%
  kable_styling(latex_options = c("striped", "scale_down"))
```

 \FloatBarrier
\newpage

## Residual Analysis 

 

 These conflicting results of students having a job between biology and the rest
 of the subjects led us to believe there might be some possible error 
 in regression. We were not inclined to believe in the existence of a linear
 relationship between our independent variables; in other terms, multicollinearity, because in all of our subjects' regressive 
 output, the F-test for overall significance coincided with the T-test for individual significance. Our next thought was to analyze the residuals' linearity and our distribution's normality. We  generated
 plots for each model that can give us insights about non-normality and heteroscedasticity. Each plot
 grouping has been labeled according to its respective subject. Consisting of two different graphs, 
 first, we plotted our Residuals vs. Fitted values, generating a line best fit 
 for our residuals; we can see if there is a linear relationship between 
 our residuals and the fitted values. For non-normality, we have plotted our 
 standardized residuals against theoretical quantities for our standard distribution. 
 This is represented as a Q-Q Plot, deviations from the one-to-one 
 linear relationship might explain where skewness may exist.

 

```{r Residual Analysis, echo = FALSE}
plot(student_regression_bio, which = c(1), main = " ")
title(main = "Biology")
plot(student_regression_bio, which = c(2))
plot(student_regression_math, which = c(1), main = " ")
title(main = "Math")
plot(student_regression_math, which = c(2))
plot(student_regression_phys, which = c(1), main = " ")
title(main = "Physics")
plot(student_regression_phys, which = c(2))
plot(student_regression_his, which = c(1), main = " ")
title(main = "History")
plot(student_regression_his, which = c(2))
```
 
 
## Results of Residual Analysis


Plotting our Standardized Residuals against our fitted values consistently
showed us a non-linear relationship. We would like to see our line of best fit 
horizontal and linear, but there seems to be some quadratic relationship
between the residuals and fitted values. This tells us that our errors are not
linear. Subsequently, these results put into question any statistical inference including 
the significance of our variables. In our Q-Q plot, we hope to have residuals hugging close
to our theoretical distribution. We see a deviation in our 
extreme theoretical values, which leads us to believe that our model departs 
from normality. Skewness in our distribution violates the assumption of 
normality in our model, which also questions the overall significance of our 
model. We cannot entirely believe that students' extracurricular activities 
are the only insignificant parameter. We must 
attempt to correct this in some capacity to gain valid results. 



```{r Log Transformation, echo = FALSE}
student_data <- student_data %>%
  mutate(
    log_math_score = log(math_score + 1),  # Adding 1 to avoid log(0)
    log_physics_score = log(physics_score + 1),
    log_history_score = log(history_score + 1),
    log_biology_score = log(biology_score + 1),
    log_absence_days = log(absence_days + 1),
    log_weekly_self_study_hours = log(weekly_self_study_hours + 1),
    extra_curricular = as.numeric(extracurricular_activities == "True"),
    has_job = as.numeric(part_time_job == "True")
  )

# Creating regression models with log-transformed variables
student_regression_math_log <- lm(log_math_score ~ log_absence_days + log_weekly_self_study_hours + extra_curricular + has_job, data = student_data)
student_regression_phys_log <- lm(log_physics_score ~ log_absence_days + log_weekly_self_study_hours + extra_curricular + has_job, data = student_data)
student_regression_his_log <- lm(log_history_score ~ log_absence_days + log_weekly_self_study_hours + extra_curricular + has_job, data = student_data)
student_regression_bio_log <- lm(log_biology_score ~ log_absence_days + log_weekly_self_study_hours + extra_curricular + has_job, data = student_data)

# Function to create a data frame of coefficients for log-transformed regression models
create_coefficients_log_dataframe <- function(regression_model, model_name) {
  # Extract the summary of the model
  model_summary <- summary(regression_model)
  
  
   # Extract coefficients and their statistics
  coefficients <- model_summary$coefficients  # This includes Estimate, Std. Error, t value, and Pr(>|t|)
  
  # Extract F-statistic and its degrees of freedom
  f_statistic <- model_summary$fstatistic["value"]
  df1 = model_summary$fstatistic["numdf"]  # Degrees of freedom for the model
  df2 = model_summary$fstatistic["dendf"]  # Degrees of freedom for the residuals
  f_p_value = pf(f_statistic, df1, df2, lower.tail = FALSE)  # Calculating the p-value for the F-statistic
  
  # Create a data frame containing the coefficients and additional model details
  df <- data.frame(
    Model = model_name,
    CoefficientValue = coefficients[, "Estimate"],
    StdError = coefficients[, "Std. Error"],
    TValue = coefficients[, "t value"],
    Pr = coefficients[, "Pr(>|t|)"],
    FStatistic = rep(f_statistic, nrow(coefficients)),
    FPValue = rep(f_p_value, nrow(coefficients)),
    ModelDF = rep(df1, nrow(coefficients)),
    ResidualDF = rep(df2, nrow(coefficients))
  )
  return(df)
}

# Assuming log-transformed models are named as student_regression_math_log, etc.
df_math_log <- create_coefficients_log_dataframe(student_regression_math_log, "Log Math")
df_phys_log <- create_coefficients_log_dataframe(student_regression_phys_log, "Log Physics")
df_his_log <- create_coefficients_log_dataframe(student_regression_his_log, "Log History")
df_bio_log <- create_coefficients_log_dataframe(student_regression_bio_log, "Log Biology")

# Combine all data frames into one
all_coefficients_log_df <- dplyr::bind_rows(df_math_log, df_phys_log, df_his_log, df_bio_log)


```

```{r Table, echo = FALSE}
#View or display the combined data frame, formatted for PDF output with LaTeX
all_coefficients_log_df %>%
 knitr::kable("latex", booktabs = TRUE, caption = "Summary Statistics for Log-Transformed Models") %>%
  kableExtra::kable_styling(latex_options = c("striped", "scale_down"))

```

\pagebreak
\FloatBarrier
## Correcting Errors in Regression
We regressed the same parameters
over a log-transformed model to correct our non-linear errors. Performing a logarithmic transformation can correct
for heteroscedasticity among errors in some cases. So, we decided to give it a 
try. 
Not only can it correct for non-linear errors, but it can reduce the skewness
that we might be dealing with due to our model facing non-normality. 
If we spread out the data clumps and combine them
for another regression, we might see different results that follow 
normality. In this case of trying to correct for non-normality and heteroscedastic 
errors, a logarithmic transformation became the apparent treatment of this model. 
Table 2, posted above, shows us the new results of our regression. If we can 
correct our errors in regression and still receive the same or similar results,
we would have much more reason to believe in the relationships of our independent 
variables on student performance. 

```{r Log Transformed Residuals, echo = FALSE}
plot(student_regression_bio_log, which = c(1), main = " ")
title(main = "Biology")
plot(student_regression_bio_log, which = c(2))
plot(student_regression_math_log, which = c(1), main = " ")
title(main = "Math")
plot(student_regression_math_log, which = c(2))
plot(student_regression_phys_log, which = c(1), main = " ")
title(main = "Physics")
plot(student_regression_phys_log, which = c(2))
plot(student_regression_his_log, which = c(1), main = " ")
title(main = "History")
plot(student_regression_his_log, which = c(2))

title(main = "History")
```

## Results of Log Transformation
Visualizing the same residual plots and comparing our non-log-transformed residual to the log-transformed, we can consistently see more linearity among our residuals. So, heteroscedasticity is 
corrected to some degree. Analyzing our Q-Q Results, we see that normality 
is corrected on the extreme positive end, but we lose normality sooner in the 
negative direction. Listed in Table 2, The significance of our model  
aligns with our original regression for math, physics, and history, 
being that extracurricular activities were the only factor that was individually
insignificant. Looking then at the relationships of each variable, we still see 
that studying is positive, absences are harmful, and having a job is also damaging. Looking at biology, we find that studying was the only significant 
factor in regression. This led us to believe the existence of other confounding
factors that were not present in our data set, such as what kind of job students
were performing, potentially students were a part of activities that were related
to the subject of study, like shadowing in a lab setting for biology, etc. 

## Conclusion

Though our confidence in our relationships 
has increased as a result of transforming our data, we cannot definitively 
claim the relationships observed are statistically significant because we 
are diverging from normality, as well as our errors are slightly 
heteroscedastic. Aside from this, we have potentially reinforced the concept of studying 
and its positive attributes on student performance, as well as potentially
identifying the relationship between how students spend their time outside of 
school and performance. Pulling quantitative and qualitative data together opened the door for 
us to see what life is like for students juggling plenty of activities. 
To further iron out and correct the errors in regression, we would like to 
observe more data from different attributes. Things like how much time students
spend with teachers outside of class, learn about their aspirations, etc. Introducing
more factors could potentially grow our understanding of these relationships, but 
may also invite more violations of regressive assumptions that need to be corrected. 
 







